Whose Truth is it Anyway? An Experiment on Annotation Bias in Times of Factual Opinion Polarization

### 2023-07-16 report/draft.Rmd
fixed references'

# Introduction
The behavior of citizens and the study thereof by social scientists both rely heavily on key information.
Classical theories of political communication, such as agenda setting or framing [e.g. @vanaelst2016political; @lecheler2019news], formulate that political information drives opinion formation and participation in politics -- from voting to protests [for overviews hereof across countries, see @pfetsch2013comparing].
The democratization of data and advent of computational social science has paved the way for new possibilities of gathering and analyzing textual data [for a recent overview, see @vanatteveldt2022computational].
In particular, advances in Natural Language Processing (NLP) have made it possible to automatically analyze large quantities of data using machine learning [e.g., see @grimmer2022text; @wei2023overview].
To determine the validity of such large scale computational analyses, we rely on "gold standard" data, created by human annotators.
It follows that the validity of these analyses hinges on the quality of these gold standards.
If a gold standard contains biases, i.e., systematic errors, it can foster bias in any downstream analysis.

@@

There is a long history of text annotation in studies analyzing political text.
While this yields lots of experiences as to how to train coders so that we get reliable hand-coded data, the procedure is expensive, protracted, and sometimes does not even get us the quality of data we need [@weber18].
While the crowd offers a solution to some of these issues [@van2021validity], the main underlying assumption that there is one correct interpretation for every input example remains untouched.
However, especially for more complex coding tasks different research fields have different interpretations of what counts as *correct* solution depending on the main goal of annotation.
Do we want to create a clean benchmark data set that follows strict linguistic rules or one that reflects how statements are being interpreted by "real" readers that do not necessarily follow theoretical definitions on which elements are needed for a specified stance?
Trends of polarization have shown that people do interpret information according to their ideological position [e.g., @rekker2022understanding; @lee2021more].
*Does this mean that some are right and others are wrong? Or is there an ideological difference in the ground truth?* These questions present a fundamental challenge to the main way of working when collecting gold standard data, as we operate from the baseline assumption that disagreement among the annotators should be avoided or reduced.
Typically, when specific cases continuously cause disagreement, more instructions are added to limit interpretations (for recent innovations to improve annotation, see for example: @barbera2021automated; @struthers2020bridging; @winter2020online; @debell2013harder; @benoit2016crowd, @ying2022topics).
However, work in computational linguistics has shown that increased annotation instructions do not increase quality [@parmar2022don].
This leaves us between a rock and a hard place.
Is there a potential bias in annotators that we should account for?

@@

Second, we looked at differences in terms of overinterpretation, which we define as coding stance in an underspecified sentence (i.e. in which the stance is not explicitly made clear) using both a strict (computational linguistics) and more lenient interpretation of what a stance is.
In other words, we tested when annotators inferred the stance of the party on the issue based on prior knowledge or beliefs about the party.
We expected that overinterpretation is more likely if the ideological distance of the coder to the party is greater (H2a) and if the coder has more political knowledge (H2b).
Overall, our findings did not support these hypotheses.
With the more lenient interpretation of overinterpretation, we did find support for H2a for the Dutch context, but we found an opposite effect for the US.
While this does not support the hypothesis in terms of the direction of the bias, it does imply that ideological distance can affect how stance is processed.
Regarding political knowledge (H2b), our results showed again country differences as well as that the way one defines a stance matters.
Our findings, for H2b in particular, pose important implications.
For sentences with clear political stances on an issue (e.g., immigration should be made more difficult) annotators generally agreed on the stance that is expressed.
But if the sentence refers to the issue without making a clear stance (e.g., many immigrants are crossing our borders) we do see a clear effect of political knowledge.
This indicates that for annotation bias to occur, there has to be sufficient room for different interpretations.
If the stance is clearly specified, then there is more of a ground truth, and the opinions and perspectives of the annotator therefore matter less.
It is when the stance is underspecified that perspectivism [@basile2021toward; @havens2022beyond] comes into play.
When a party talks about a political issue without taking an explicit stance, annotators need to fill in the gaps and make their own inferences about the party's stance on this issue. Considering that in many political documents and specifically reporting on political news statements are not made explicit, underspecified sentences can be assumed to rather be the rule than the exception.
Our results therefore imply that political knowledge is indeed a relevant dimension in shaping how an annotator perceives political stances.

>>>>

# Introduction
The behavior of citizens and the study thereof by social scientists both rely heavily on key information.
Classical theories of political communication, such as agenda setting or framing [e.g. @vanaelst2016political; @lecheler2019news], formulate that political information drives opinion formation and participation in politics -- from voting to protests [for overviews hereof across countries, see @pfetsch2013comparing].
The democratization of data and advent of computational social science has paved the way for new possibilities of gathering and analyzing textual data [for a recent overview, see @vanatteveldt2022computational].
In particular, advances in Natural Language Processing (NLP) have made it possible to automatically analyze large quantities of data using machine learning [e.g., see @grimmer2022text; @akyurek-etal-2020-multi, @piskorski2023semeval].
To determine the validity of such large scale computational analyses, we rely on "gold standard" data, created by human annotators.
It follows that the validity of these analyses hinges on the quality of these gold standards.
If a gold standard contains biases, i.e., systematic errors, it can foster bias in any downstream analysis.

@@

There is a long history of text annotation in studies analyzing political text.
While this yields lots of experiences as to how to train coders so that we get reliable hand-coded data, the procedure is expensive, protracted, and sometimes does not even get us the quality of data we need [@weber18].
While the crowd offers a solution to some of these issues [@van2021validity], the main underlying assumption that there is one correct interpretation for every input example remains untouched.
However, especially for more complex coding tasks different research fields have different interpretations of what counts as *correct* solution depending on the main goal of annotation.
Do we want to create a clean benchmark data set that follows strict linguistic rules or one that reflects how statements are being interpreted by "real" readers that do not necessarily follow theoretical definitions on which elements are needed for a specified stance?
Trends of polarization have shown that people do interpret information according to their ideological position [e.g., @rekker2022understanding; @lee2021more].
*Does this mean that some are right and others are wrong? Or is there an ideological difference in the ground truth?* These questions present a fundamental challenge to the main way of working when collecting gold standard data, as we operate from the baseline assumption that disagreement among the annotators should be avoided or reduced.
Typically, when specific cases continuously cause disagreement, more instructions are added to limit interpretations (for recent innovations to improve annotation, see for example: @barbera2021automated; @struthers2020bridging; @winter2020online; @debell2013harder; @benoit2016crowd, @ying2022topics).
However, work in computational linguistics has shown that increased annotation instructions do not increase quality [@parmar2023don].
This leaves us between a rock and a hard place.
Is there a potential bias in annotators that we should account for?

@@

Second, we looked at differences in terms of overinterpretation, which we define as coding stance in an underspecified sentence (i.e. in which the stance is not explicitly made clear) using both a strict (computational linguistics) and more lenient interpretation of what a stance is.
In other words, we tested when annotators inferred the stance of the party on the issue based on prior knowledge or beliefs about the party.
We expected that overinterpretation is more likely if the ideological distance of the coder to the party is greater (H2a) and if the coder has more political knowledge (H2b).
Overall, our findings did not support these hypotheses.
With the more lenient interpretation of overinterpretation, we did find support for H2a for the Dutch context, but we found an opposite effect for the US.
While this does not support the hypothesis in terms of the direction of the bias, it does imply that ideological distance can affect how stance is processed.
Regarding political knowledge (H2b), our results showed again country differences as well as that the way one defines a stance matters.
Our findings, for H2b in particular, pose important implications.
For sentences with clear political stances on an issue (e.g., immigration should be made more difficult) annotators generally agreed on the stance that is expressed.
But if the sentence refers to the issue without making a clear stance (e.g., many immigrants are crossing our borders) we do see a clear effect of political knowledge.
This indicates that for annotation bias to occur, there has to be sufficient room for different interpretations.
If the stance is clearly specified, then there is more of a ground truth, and the opinions and perspectives of the annotator therefore matter less.
It is when the stance is underspecified that perspectivism [@cabitza2023toward; @havens2022beyond] comes into play.
When a party talks about a political issue without taking an explicit stance, annotators need to fill in the gaps and make their own inferences about the party's stance on this issue. Considering that in many political documents and specifically reporting on political news statements are not made explicit, underspecified sentences can be assumed to rather be the rule than the exception.
Our results therefore imply that political knowledge is indeed a relevant dimension in shaping how an annotator perceives political stances.
